### 2018-07-02 paper/1-intro.tex
Small edits from simon

Build systems (such as \Make) are big, complicated, and used by every
software developer on the planet.  But they are a sadly unloved part
of the software ecosystem, very much a means to an end, and seldom the
focus of attention.
% Rarely do people ask questions like ``What does it mean for my build
% system to be correct?'' or ``What are the trade-offs between different
% approaches?''.  Moreover,
Complex build systems use subtle algorithms, but they
are often hidden away, and not the object of study.
Recently, the challenges of scale have driven large software firms
like Microsoft, Facebook, and Google to develop their own build
systems, exploring new points in the design space.

In this paper we offer a general framework in which to understand
and compare build systems,
in a way that is both abstract (omitting incidental detail)
and yet precise (implemented as Haskell code:
\url{https://hackage.haskell.org/package/build-0.0.1.1}). Specifically we make
these contributions:
\begin{itemize}
\item Build systems vary on many axes, including: static vs dynamic
  dependencies; local vs cloud; deterministic vs non-deterministic build rules;
  support for early cutoff; self-tracking build systems; and the type of
  persistent build information. In~\S\ref{sec-background} we identify some key
  properties, in the context of four carefully-chosen build systems.

@@

\end{itemize}
In short, instead of seeing build systems as unrelated
points in space, we now see them as locations in a landscape,
leading to a better understanding of what they do and how they compare,
and suggesting exploration of other (as yet unoccupied points) in the
landscape.
We discuss engineering aspects in \S\ref{sec-engineering}, and related
work in \S\ref{sec-related}.

>>>>

Build systems (such as \Make) are big, complicated, and used by every
software developer on the planet.  But they are a sadly unloved part
of the software ecosystem, very much a means to an end, and seldom the
focus of attention.
% Rarely do people ask questions like ``What does it mean for my build
% system to be correct?'' or ``What are the trade-offs between different
% approaches?''.  Moreover,
For years \Make{} dominated, but more recently
Recently, the challenges of scale have driven large software firms
like Microsoft, Facebook, and Google to develop their own build
systems, exploring new points in the design space.
These complex build systems use subtle algorithms, but they
are often hidden away, and not the object of study.

In this paper we offer a general framework in which to understand
and compare build systems,
in a way that is both abstract (omitting incidental detail)
and yet precise (implemented as Haskell code:
\url{https://hackage.haskell.org/package/build-0.0.1.1}). Specifically we make
these contributions:
\begin{itemize}
\item Build systems vary on many axes, including: static vs dynamic
  dependencies; local vs cloud; deterministic vs non-deterministic build rules;
  support for early cutoff; self-tracking build systems; and the type of
  persistent build information. In~\S\ref{sec-background} we identify some key
  properties, illustrated by four carefully-chosen build systems.

@@

\end{itemize}
In short, instead of seeing build systems as unrelated
points in space, we now see them as locations in a connected landscape,
leading to a better understanding of what they do and how they compare,
and suggesting exploration of other (as yet unoccupied points) in the
landscape.
We discuss engineering aspects in \S\ref{sec-engineering}, and related
work in \S\ref{sec-related}.

### 2018-07-02 paper/2-background.tex
Small edits from simon

Unlike \Make, \Excel does not need to know all task dependencies upfront. Some
dependencies may change \emph{dynamically} according to computation results. For
example:

@@

To support dynamic dependencies, \Excel's build engine~\cite{excel_recalc} is
significantly different from \Make. \Excel arranges the cells into a linear
sequence, called the \emph{calc chain}.  During the build, \Excel processes
cells in the calc-chain sequence, but if computing a cell \cmd{C} requires the
value of a cell \cmd{D} that has not yet been computed, \Excel \emph{aborts}
computation of \cmd{C}, moves \cmd{D} before \cmd{C} in the calc chain, and
resumes the build starting with \cmd{D}. When a build is complete, the resulting
calc chain respects all the dynamic dependencies of the spreadsheet. When an
input value (or formula) is changed, \Excel uses the final calc chain from the
\emph{previous} build as its starting point so that, in the common case where
changing an input value does not change dependencies, there are no aborts.
Notice that build always succeeds regardless of the initial calc chain (barring
truly circular dependencies); the calc chain is just an optimisation.
We refer to this algorithm as \emph{restarting}, and will discuss it
in more detail in~\S\ref{sec-implementation-excel}.

Dynamic dependencies complicate minimality.  In the above example, \cmd{B1}
should only be recomputed if \cmd{A1} or \cmd{C1} change, but not if (say)
\cmd{A2} changes; but these facts are not statically apparent. In practice
\Excel implements a conservative approximation to minimality: it recomputes a
formula if~(i)~the formula mentions a changed cell, or~(ii)~the formula uses a
function like \cmd{INDIRECT} whose dependencies are not statically visible,
or~(iii)~the formula itself has changed.

Item (iii) in the above list highlights a distinguishing feature of \Excel: it
is \emph{self-tracking}. Most build systems only track changes of inputs and
intermediate results, but \Excel also tracks changes in the tasks themselves: if
a formula is modified, \Excel will recompute it and propagate the changes.
Self-tracking is uncommon in software build systems, where one often needs to
manually initiate a full rebuild even if just a single build task has changed.
We discuss self-tracking further in~\S\ref{sec-tracking-aspects}.

@@

\Bazel is one of the first examples of openly available cloud build systems.
As of writing, it is not possible to express dynamic dependencies in
user-defined build rules, however some of the predefined build rules require
dynamic dependencies and the internal build engine can cope with them by using
a \emph{restarting} task scheduler, which is similar to that of \Excel but does
not use the calc chain. \Bazel is not minimal in the sense that it may restart a
task multiple times as new dependencies are discovered and rebuilt, but it
supports the early cutoff optimisation.

>>>>

Unlike \Make, \Excel does not need to know all task dependencies upfront. Indeed, some
dependencies may change \emph{dynamically} according to computation results. For
example:

@@

To support dynamic dependencies, \Excel's calc engine~\cite{excel_recalc} is
significantly different from \Make. \Excel arranges the cells into a linear
sequence, called the \emph{calc chain}.  During the build, \Excel processes
cells in the calc-chain sequence, but if computing a cell \cmd{C} requires the
value of a cell \cmd{D} that has not yet been computed, \Excel \emph{aborts}
computation of \cmd{C}, moves \cmd{D} before \cmd{C} in the calc chain, and
resumes the build starting with \cmd{D}. When a build is complete, the resulting
calc chain respects all the dynamic dependencies of the spreadsheet. When an
input value (or formula) is changed, \Excel uses the final calc chain from the
\emph{previous} build as its starting point so that, in the common case where
changing an input value does not change dependencies, there are no aborts.
Notice that build always succeeds regardless of the initial calc chain (barring
truly circular dependencies); the calc chain is just an optimisation.
We refer to this algorithm as \emph{restarting}, and will discuss it
in more detail in~\S\ref{sec-implementation-excel}.

Dynamic dependencies complicate minimality.  In the above example, \cmd{B1}
should only be recomputed if \cmd{A1} or \cmd{C1} change, but not if (say)
\cmd{A2} changes; but these facts are not statically apparent. In practice
\Excel implements a conservative approximation to minimality: it recomputes a
formula if~(i)~the formula statically mentions a changed cell, or~(ii)~the formula uses a
function like \cmd{INDIRECT} whose dependencies are not statically visible,
or~(iii)~the formula itself has changed.

Item (iii) in the above list highlights another distinguishing feature of \Excel: it
is \emph{self-tracking}. Most build systems only track changes of inputs and
intermediate results, but \Excel also tracks changes in the tasks themselves: if
a formula is modified, \Excel will recompute it and propagate the changes.
Self-tracking is uncommon in software build systems, where one often needs to
manually initiate a full rebuild even if just a single build task has changed.
We discuss self-tracking further in~\S\ref{sec-tracking-aspects}.

@@

\Bazel is one of the first openly-available cloud build systems.
As of writing, it is not possible to express dynamic dependencies in
user-defined build rules; however some of the pre-defined build rules require
dynamic dependencies and the internal build engine can cope with them by using
a \emph{restarting} task scheduler, which is similar to that of \Excel but does
not use the calc chain. \Bazel is not minimal in the sense that it may restart a
task multiple times as new dependencies are discovered and rebuilt, but it
supports the early cutoff optimisation.

### 2018-07-02 paper/3-abstractions.tex
Small edits from simon

\noindent
Here \hs{Task} describes a single build task, while \hs{Tasks} associates a
task description to every non-input key; input keys are associated with
\hs{Nothing}. The highly-abstracted type\footnote{Readers familiar with
\emph{lenses} or \emph{profunctor optics} might recognise a familiar pattern.
We discuss this in~\S\ref{sec-related-optics}.} \hs{Task} is best introduced by
an example. Consider this \Excel spreadsheet:

>>>>

\noindent
A \hs{Task} describes a single build task, while \hs{Tasks} associates a
\hs{Task} to every non-input key; input keys are associated with
\hs{Nothing}. The highly-abstracted type\footnote{Readers familiar with
\emph{lenses} or \emph{profunctor optics} might recognise a familiar pattern.
We discuss this in~\S\ref{sec-related-optics}.} \hs{Task} is best introduced by
an example. Consider this \Excel spreadsheet:

### 2018-07-02 paper/4-build.tex
Add an extra line
Deep traces don't support early cutoff
Small edits from simon

The build systems overview (\S\ref{sec-background-summary}) highlighted three
distinct task schedulers respecting the dependency order. This subsection
explores their properties and possible implementations.

@@

The topological scheduler pre-computes a linear order, which when followed,
ensures the build result is correct regardless of the initial store. Given a
task description and the output \hs{key}, you can compute the linear order by
first finding the (acyclic) graph of the \hs{key}'s reachable dependencies, and
then computing a topological sort. However, as we have seen in~\S\ref{sec-deps},
we can only extract dependencies from an applicative task, which requires the
build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling out dynamic
dependencies.

@@

The topological scheduler has two downsides: it is limited to \hs{Applicative}
build systems and requires a fresh topological sort each time. So, while the
actions themselves may be incremental (i.e. unnecessary tasks will not be
performed), the pre-processing is not. We can try to incrementalise the
topological sort by storing the topological order between build runs and assume
it to be correct, fixing it up if the build discovers any inconsistency, as
\Excel does.

@@

The second aspect, determining what to rebuild, can be addressed in one of four
fundamental ways, with a number of tweaks and variations within them.

@@

\noindent
We assume that \hs{Hash}~\hs{v} is a small constant size, constructed from
hashing the underlying \hs{v} rather than storing it directly. After
successfully building a key, we call \hs{recordVT} passing the key, the hash of
its value, and the list of hashes and dependencies. Combined with the existing
set of verifying traces, that provides a new set of verifying traces.

More interestingly, to \emph{verify} whether a key needs rebuilding we supply
the key, the hash of its current value, a function for obtaining the post-build
value of any key (using a scheduling strategy as per
\S\ref{sec-dependency-orderings}), and the existing \hs{VT} information. The
result will be a \hs{Bool} where \hs{True} indicates that the current value is
up to date, and \hs{False} indicates that it should be rebuilt.

One potential implementation would be to record all arguments passed to
\hs{recordVT} in a list, and verify by simply checking if any list item matches
the information passed by \hs{verifyVT} (we discuss smarter implementations
in~\S\ref{sec-smart-traces}).

@@

A verifying trace allows us to mark a key dirty and rebuild it. Extending that
information we can store a \emph{constructive} trace which is the trace plus the
actual resulting value. Once we are storing the complete result it makes sense
to record many constructive traces per key, and to share them with other users,
providing cloud-build functionality. We can represent this additional
information by providing the operations:

@@

\noindent
The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
the hash of the resulting value, we require the whole value. The \hs{verifyVT}
has been replaced with \hs{constructCT}, which instead of taking the hash of the
value, returns a list of suitable values. If the current value in the store
matches one of the possible values, the build can skip this rule. If the
resulting list is empty, the key must be rebuilt. However, if the current value
does not match the store, and there is a possible value, we can use any value
from the constructive list \emph{without} doing any work to build it.

@@

Constructive traces always verify keys by looking at their immediate
dependencies, which must have first been brought up to date, meaning that the
time to verify a key is based on the number of transitive dependencies. A
\emph{deep} constructive trace optimises this process by only looking at the
terminal input keys, ignoring any intermediate dependencies. The operations
capturing this approach are the same as for constructive traces
in~\S\ref{sec-constructive-traces}, but we use the names \hs{recordDCT} and
\hs{constructDCT}, where the underlying \hs{DCT} representation need only record
information about hashes of inputs, not intermediate dependencies.

The deep constructive trace approach only works if the operations are
\emph{deterministic}, otherwise inconsistencies arise
(see~\S\ref{sec-non-determinism}). Concretely, if the result of building a key
given a set of immediate dependencies can vary, then it is possible that key
\hs{k} will have been computed locally, but something \emph{using} \hs{k} will
have been taken as a constructive input using a different value for \hs{k}. If
another rule combines those two values then two different values of \hs{k} will
have been used in one rule, leading to problems such as link-errors for typical
compilers.

While current build systems using deep constructive traces store only hashes of
inputs, that is not a strict requirement, as the technique works equally well by
skipping $n$ dependency levels, where the input-only approach is the special
case of $n = \infty$. Similarly, constructive traces can be described as taking
the direct dependencies, and thus having depth $n = 1$. A downside of deep
constructive traces is that they cannot support cutoff
(\S\ref{sec-background-shake}), other than at the $n$ levels of dependencies. On
the other hand, these traces are particularly useful for ``shallow builds'', as
discussed in~\S\ref{sec-cloud-aspects}.

@@

\begin{enumerate}
\item Any system using verifying traces, e.g. \Shake, is unlikely to see
significant benefit from storing more than one \hs{Trace} per key\footnote{There
is a small chance of a benefit if the dependencies change but the result does
not, and then the dependencies change back to what they were before.}.
Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v)},
where the initial \hs{k} is the \hs{key} field of \hs{Trace}.

\item Any system using \hs{Applicative} dependencies can omit the dependency
keys from the \hs{Trace} as they can be recovered from the \hs{key} field.

\item Any \hs{Applicative} build system storing constructive traces, e.g.
\Bazel, can index directly from the key and results to the output result~--~i.e.
\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
are stored on a central server, the client can compute the key and the hashes of
its dependencies, then make a single call to the server to retrieve the result.

@@

\begin{table}
\vspace{-1mm}
\smaller
\centering
\begin{tabular}{l||c|c|c}
\hline
Property           & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
\hline
Dirty bit\hfill\S\ref{sec-dirty-bit}                                                             & \Make       & \Excel & -              \\\hline
Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\\hline
Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\\hline
Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
\end{tabular}
\vspace{1mm}
\caption{Build systems \`a la carte.\label{tab-build-systems}}
\vspace{-8mm}
\end{table}

With the information in this section we can build a table comparing the
dependency order strategy with the out-of-date keys strategy, providing 12~possible
build systems, 8~of which are actually inhabited by existing build systems
(we discuss all these systems in \S\ref{sec-related-build}). Of the
remaining 4~spots, all result in workable build systems. The most interesting
unfilled spot in the table is suspending constructive traces, which would provide
many benefits, and which we title \Cloud \Shake (as we plan on extending \Shake to occupy that spot)
and further explore further in~\S\ref{sec-implementation-cloud}.

>>>>

Section \S\ref{sec-background} introduced three different \emph{task scehedulers}
that decide which tasks to execute and in what order; see the ``Scheduler'' column
of Table~\ref{tab-summary} in \S\ref{sec-background-summary}.
This subsection explores the properties of the three schedulers, and
possible implementations.

@@

The topological scheduler pre-computes a linear order of tasks, which when followed,
ensures the build result is correct regardless of the initial store. Given a
task description and the output \hs{key}, you can compute the linear order by
first finding the (acyclic) graph of the \hs{key}'s reachable dependencies, and
then computing a topological sort. However, as we have seen in~\S\ref{sec-deps},
we can only extract dependencies from an applicative task, which requires the
build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling out dynamic
dependencies.

@@

The topological scheduler has two downsides: it is limited to \hs{Applicative}
build systems and requires a fresh topological sort each time. So, while the
actions themselves may be incremental (i.e. unnecessary tasks will not be
performed), the pre-processing is not. We can try to incrementalise the
topological sort by storing the topological order between build runs,
using it as a starting point in the next run, 
fixing it up if the build discovers any inconsistency, as
\Excel does.

@@

Suppose the scheduler decides that that a key \hs{k} should be
brought up to date.  The next question is: does any work need to be done,
or is it already up to date?   Or, in a cloud-build system, do we have
a cached copy of the value we need?
These questions can be addressed in one of four
fundamental ways, with a number of tweaks and variations within them.

@@

\noindent
Rather than storing (large) values \hs{v}, the verifying trace \hs{VT}
stores only hashes, of type \hs{Hash}~\hs{v}, of those values.  Since the
verfying trace persists from one build to the next -- it constitutes the build system's ``memory''
-- it is helpful for it to be of modest size.
After successfully building a key, we call \hs{recordVT} to add a record
to the current \hs{VT}, passing the key, the hash of
its value, and the list of hashes and dependencies.

More interestingly, to \emph{verify} whether a key needs rebuilding we supply
the key, the hash of its current value, a function for obtaining the post-build
value of any key (using a scheduling strategy as per
\S\ref{sec-dependency-orderings}), and the existing \hs{VT} information. The
result will be a \hs{Bool} where \hs{True} indicates that the current value is
already up to date, and \hs{False} indicates that it should be rebuilt.

One potential implementation would be to record all arguments passed to
\hs{recordVT} in a list, and verify by simply checking if any list item matches
the information passed by \hs{verifyVT}.  We discuss smarter implementations
in~\S\ref{sec-smart-traces}.

A verifying trace, and other types of traces discussed in this section, support
dynamic dependencies and minimality; furthermore, all traces except for deep
traces~(\S\ref{sec-deep-constructive-traces}) support early cutoff.

@@

A verifying trace deliberately records only small hashes, so that it can be small.
A \emph{constructive} trace additionally stores the resulting value.
Once we are storing the complete result it makes sense
to record many constructive traces per key, and to share them with other users,
providing cloud-build functionality. We can represent this additional
information by providing the operations:

@@

\noindent
The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
the hash of the resulting value, we require the whole value. The \hs{verifyVT}
has been replaced with \hs{constructCT}, which instead of taking the hash of the
current value as \emph{input}, returns a list of suitable values as \emph{output}.
If the current value in the store
matches one of the possible values, the build can skip this rule. If the
resulting list is empty, the key must be rebuilt. However, if the current value
does not match the store, and there is a possible value, we can use any value
from the constructive list \emph{without} doing any work to build it, and copy
it into the store from the cloud.

@@

Constructive traces always verify keys by looking at their immediate
dependencies, which must have first been brought up to date, meaning that the
time to verify a key is based on the number of transitive dependencies. A
\emph{deep} constructive trace optimises this process by only looking at the
terminal \emph{input keys}, ignoring any intermediate dependencies. The operations
capturing this approach are the same as for constructive traces
in~\S\ref{sec-constructive-traces}, but we use the names \hs{recordDCT} and
\hs{constructDCT}, where the underlying \hs{DCT} representation need only record
information about hashes of inputs, not intermediate dependencies.

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well
if we skip any number dependency levels (say $n$ levels).
The input-only approach is the special
case of $n = \infty$, and constructive trances are the special case of 
$n = 1$.

The deep constructive trace approach only works if the operations are
\emph{deterministic}, otherwise inconsistencies arise
(see~\S\ref{sec-non-determinism}). \simon{That section doesn't actually
  mention ``inconsistencies''.
  Also I think the inconsistencies are called Frankenbuilds, a term we might
  want to introduce.}
Concretely, if the result of building a key
given a set of immediate dependencies can vary, then it is possible that key
\hs{k} will have been computed locally, but something \emph{using} \hs{k} will
have been taken as a constructive input using a different value for \hs{k}. If
another rule combines those two values then two different values of \hs{k} will
have been used in one rule, leading to problems such as link-errors for typical
compilers.

A downside of deep constructive traces is that they cannot support cutoff
(\S\ref{sec-background-shake}), other than at the $n$ levels of dependencies. On
the other hand, these traces are particularly useful for ``shallow builds'', as
discussed in~\S\ref{sec-cloud-aspects}.

@@

\begin{enumerate}
\item Any system using verifying traces, e.g. \Shake, is unlikely to see
significant benefit from storing more than one \hs{Trace} per key\footnote{There
is a small chance of a benefit if the dependencies change but the result does
not, and then the dependencies change back to what they were before.}.
Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v)},
where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
\simon{The data type decl had three args, k, v, r.}

\item Any system using \hs{Applicative} dependencies can omit the dependency
keys from the \hs{Trace} since they can be recovered from the \hs{key} field.

\item Any \hs{Applicative} build system using constructive traces, e.g.
\Bazel, can index directly from the key and results to the output result~--~i.e.
\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
are stored on a central server, the client can compute the key and the hashes of
its dependencies, then make a single call to the server to retrieve the result.

@@

\begin{table}
\vspace{-1mm}
\smaller
\centering
\begin{tabular}{l||c|c|c}
\hline
 & \multicolumn{3}{c}{Scheduling strategy} \\
Out-of-date key strategy  & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
\hline
Dirty bit\hfill\S\ref{sec-dirty-bit}                                                             & \Make       & \Excel & -              \\\hline
Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\\hline
Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\\hline
Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
\end{tabular}
\vspace{1mm}
\caption{Build systems \`a la carte.\label{tab-build-systems}}
\vspace{-8mm}
\end{table}

With the information in this section we can build Table~\ref{tab-build-systems}, which
tabulates combinations of the schedling strategy and the out-of-date keys strategy,
providing 12~possible build systems, 8~of which are actually inhabited by existing build systems
(we discuss all these systems in \S\ref{sec-related-build}). Of the
remaining 4~spots, all result in workable build systems. The most interesting
unfilled spot in the table is suspending constructive traces, which would provide
many benefits, and which we title \Cloud \Shake
% (as we plan on extending \Shake to occupy that spot)
and further explore further in~\S\ref{sec-implementation-cloud}.

### 2018-07-02 paper/5-implementations.tex
Small edits from simon

In the previous sections we discussed the types of build systems, and how they
can be broken down into two main components: a scheduler and a rebuilder.
In this section we make this abstract distinction concrete, by
implementing a number of build systems as a composition of a scheduler
and a rebuilder.  Our design is expressed by these types:
\begin{minted}[xleftmargin=10pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
So a \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
it to construct a \hs{Build} system, by choosing which keys to rebuild in which
order. The \hs{Rebuilder} makes uses of the persistent build information
\hs{ir}, while the scheduler might augment that with further persistent
information of its own, yielding \hs{i}.

@@

The \hs{modTimeRebuilder} uses the pair
\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
build information, carried by a \hs{MonadState} monad. This
\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
modification times}. The rebuilder receives three arguments: a
\hs{key}, its current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
the \hs{key} if necessary.  The rebuilder first decides if the
\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the \hs{key} is not
found, that must mean it has never been built before; otherwise
\hs{modTimeRebuilder} can see if any of the \hs{task}'s dependencies
(computed by \hs{dependencies}) are out of date. If the \hs{key} is
\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
the new modification time of the \hs{key}\footnote{The real \Make
relies on the file system to track modification times, but we prefer
to make it explicit here.}; otherwise we can just return the current \hs{value}.
We assume that the external system, which invokes the build system, updates
\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.

@@

The \hs{dirtyBitRebuilder} is very simple: if the \hs{key}'s dirty bit is set,
we \hs{run} the \hs{task} to rebuild the \hs{key}; otherwise we return the
current \hs{value} as is. The external system, which invokes \Excel's build
engine, is required to provide a transitively closed set of dirty bits.
Therefore, unlike \Make's \hs{modTimeRebuilder}, the \hs{dirtyBitRebuilder} does
not need to modify \hs{i} to trigger rebuilds of dependent keys.

>>>>

In the previous sections we discussed the types of build systems, and how they
can be broken down into two main components: a scheduler and a rebuilder.
In this section we make this abstract distinction concrete, by
implementing a number of build systems as a composition of a scheduler
and a rebuilder.  Our design is expressed by these types (Figure~\ref{fig-types})
\begin{minted}[xleftmargin=10pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
A \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
it to construct a \hs{Build} system, by choosing which keys to rebuild in which
order. The \hs{Rebuilder} makes uses of the persistent build information
\hs{ir}, while the scheduler might augment that with further persistent
information of its own, yielding \hs{i}.

@@

The \hs{modTimeRebuilder} uses the pair
\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
build information, carried by a \hs{MonadState} monad. This
\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
modification times}.
We assume that the external system, which invokes the build system, updates
\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.

The rebuilder receives three arguments: a
\hs{key}, its current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
the \hs{key} if necessary.  The rebuilder first decides if the
\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the \hs{key} is not
found, that must mean it has never been built before; otherwise
\hs{modTimeRebuilder} can see if any of the \hs{task}'s dependencies
(computed by \hs{dependencies}) are out of date. If the \hs{key} is
\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
the new modification time of the \hs{key}\footnote{The real \Make
relies on the file system to track modification times, but we prefer
to make it explicit here.}; otherwise we can just return the current \hs{value}.

@@

The external system, which invokes \Excel's build engine, is required
to provide a transitively closed set of dirty bits.  That is, if a
cell is changed, its dirty bit is set, and the dirty bit of any cell
whose value might perhaps change as a result.  It is OK to mark too many
cells as dirty; but not OK to mark too few.

The \hs{dirtyBitRebuilder} is very simple: if the \hs{key}'s dirty bit is set,
we \hs{run} the \hs{task} to rebuild the \hs{key}; otherwise we return the
current \hs{value} as is.
Because the dirty cells are transitively closed,
unlike \Make's \hs{modTimeRebuilder}, the \hs{dirtyBitRebuilder} does
not need to modify \hs{i} to trigger rebuilds of dependent keys.

### 2018-07-02 paper/6-engineering.tex
Small edits from simon

While we have given simple implementations assuming a single thread of execution,
all the build systems we address can actually build independent keys in parallel.
While it complicates the model, the complications can be restricted exclusively
to the scheduler:

@@

\begin{description}
\item[Untracked state] Some tasks depend on untracked state -- for example
C compilation will explicitly list the \cmd{source.c} file as a dependency,
but it may not record that the version of \cmd{gcc} is also a dependency.

@@

\item[Volatility] Some tasks are defined to change in every execution, for
example \Excel provides a~``function'' \cmd{RANDBETWEEN} which produces a random
number in a specified range~--~on each recalculation it is defined to change.
Similarly, build systems like \Make and \Shake provide \emph{phony rules} which
are also volatile.
\end{description}

Interestingly, there is significant interplay between all three sources of
impurity. Systems like \Bazel use various sandboxing techniques to guard against
missing dependencies, but none are likely to capture all dependencies right down
to CPU model and microcode version. Rules that do have untracked state can be
marked as volatile, a technique \Excel takes with the \cmd{INDIRECT} function,
removing the untracked state at the cost of minimality.

@@

\item[Frankenbuilds] A build is considered a
\emph{frankenbuild}~\cite{esfahani2016cloudbuild} if a value is calculated
locally, but something that depends on that key is pulled from the cache, and
the value calculated locally does not match what was previously calculated and
stored in the cloud. Our implementations avoid this issue by storing complete
traces, but if a cloud build system was to only reference input nodes this
situation can arise.

@@

\noindent
Namely, first look up the formula, then interpret it. It is not possible to
change dependencies based on the formula in a \hs{Tasks}~\hs{Applicative}, as
would be required, so instead the formula can be captured as a dependency
(but its value not used) and \emph{also} baked directly into the \hs{sprsh5}
function.

The build systems that have precise self-tracking are all ones which use a
\emph{non-embedded domain specific language} to describe computations.
Those which make use of a full programming language, e.g. \Shake, are faced with
the challenge of implementing equality on arbitrary task functions. For such
build systems the incredibly pessimistic assumption of saying that any change to
the build system potentially changes any build rule can often be used~--~the
classic example being a makefile depending on itself.

>>>>

We have given simple implementations assuming a single thread of execution,
but all the build systems we address can actually build independent keys in parallel.
While it complicates the model, the complications can be restricted exclusively
to the scheduler:

@@

\begin{description}
\item[Untracked dependencies] Some tasks depend on untracked values -- for example
C compilation will explicitly list the \cmd{source.c} file as a dependency,
but it may not record that the version of \cmd{gcc} is also a dependency.

@@

\item[Volatility] Some tasks are defined to change in every execution, for
example \Excel provides a~``function'' \cmd{RANDBETWEEN} which produces a random
number in a specified range~--~on each recalculation it is defined to change.
Similarly, build systems like \Make and \Shake provide \emph{phony rules} which
are also volatile.
\end{description}
\simon{Is there a difference between non-determinism and volatility?  Perhaps
  one is accidental non-det while volatility is by-design non-det? But does the
  distinction matter?   Maybe for accidental non-det you are free to use an
  earlier cached result, whereas for volatity you should always recompute?
  Volatility is best handled by a dependency on RealWorld, adn we coudl say that right here.}

Interestingly, there is significant interplay between all three sources of
impurity. Systems like \Bazel use various sandboxing techniques to guard against
missing dependencies, but none are likely to capture all dependencies right down
to CPU model and microcode version. Rules that do have untracked dependencies can be
marked as volatile, a technique \Excel takes with the \cmd{INDIRECT} function,
removing the untracked dependency at the cost of minimality.

@@

\item[Frankenbuilds] A build is considered a
\emph{frankenbuild}~\cite{esfahani2016cloudbuild} if a value is calculated
locally, but something that depends on that key is pulled from the cache, and
the value calculated locally does not match what was previously calculated and
stored in the cloud. \simon{Don't understand. The explanation in 4.2.4 is better -- and I think describes
the same thing?}
Our implementations avoid this issue by storing complete
traces, but if a cloud build system was to only reference input nodes this
situation can arise.

@@

\noindent
Namely, first look up the formula, then interpret it. It is not possible to
change dependencies based on the formula in a \hs{Tasks}~\hs{Applicative}, as
would be required, so instead the formula can be captured as a dependency
(but its value not used) and \emph{also} baked directly into the \hs{sprsh5}
function. \simon{Don't understand.}

The build systems that have precise self-tracking are all ones which use a
\emph{non-embedded domain specific language} to describe computations.
\simon{I think you mean that you can find dependencies with executing the function,
  which seems quite difference from ``backing it into'' the Haskell code.}
Those which make use of a full programming language, e.g. \Shake, are faced with
the challenge of implementing equality on arbitrary task functions. For such
build systems the incredibly pessimistic assumption of saying that any change to
the build system potentially changes any build rule can often be used~--~the
classic example being a makefile depending on itself.
